---
title: "hipercow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{hipercow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
source("common.R")
vignette_root <- new_hipercow_root_path()
fs::file_copy("simulation.R", vignette_root)
set_vignette_root(vignette_root)
cleanup <- withr::with_dir(
  vignette_root,
  hipercow::hipercow_example_helper(with_logging = TRUE,
                                    new_directory = FALSE,
                                    initialise = FALSE))
```

Parallel computing on a cluster can be more challenging than running things locally because it's often the first time that you need to package up code to run elsewhere, and when things go wrong it's more difficult to get information on why things failed.

Much of the difficulty of getting things running involves working out what your code depends on, and getting that installed in the right place on a computer that you can't physically poke at.  The next set of problems is dealing with the ballooning set of files that end up being created - templates, scripts, output files, etc.

The `hipercow` package aims to remove some of this pain, with the aim that running a task on the cluster should be (almost) as straightforward as running things locally, at least once some basic setup is done.

At the moment, this document assumes that we will be using the "Windows" cluster, which implies the existence of some future non-Windows cluster. Stay tuned.

This manual is structured in escalating complexity, following the chain of things that a hypothetical user might encounter as they move from their first steps on the cluster through to running enormous batches of tasks.

# Installing prerequisites

Install the required packages from our "r-universe". Be sure to run this in a fresh session.

```r
install.packages(
  "hipercow",
  repos = c("https://mrc-ide.r-universe.dev", "https://cloud.r-project.org"))
```

Once installed you can load the package with

```{r}
library(hipercow)
```

or use the package by prefixing the calls below with `hipercow::`, as you prefer.

Follow any platform-specific instructions in `vignettes("<cluster>")`; this will depend on the cluster you intend to use:

* Windows: `vignette("windows")`

# Filesystems and paths

We need a concept of a "root"; the point in the filesystem we can think of everything relative to.  This will feel familiar to you if you have used git or orderly, as these all have a root (and this root will be a fine place to put your cluster work). Typically all paths will be *within* this root directory, and paths above it, or absolute paths in general, effectively cease to exist. If your project works this way then it's easy to move around, which is exactly what we need to do in order to run it on the cluster.

If you are using RStudio, then we strongly recommend using an [RStudio project](https://support.posit.co/hc/en-us/articles/200526207-Using-RStudio-Projects).

# Initialising

Run

```{r}
hipercow_init()
```

which will write things to a new path `hipercow/` within your working directory.

After initialisation you will typically want to configure a "driver", which controls how tasks are sent to clusters.  At the moment the only option is the windows cluster so for practical work you would write:

```r
hipercow_configure(driver = "windows")
```

however, for this vignette we will use a special "example" driver which simulates what the cluster will do (don't use this for anything yourself, it really won't help):

```{r}
hipercow_configure(driver = "example")
```

You can run initialisation and configuration in one step by running

```r
hipercow_init(driver = "windows")
```

After initialisation and configuration you can see the computed configuration by running `hipercow_configuration()`:

```{r}
hipercow_configuration()
```

Here, you can see versions of important packages, information about where you are working, and information about how you intend to interact with the cluster.  See `vignette("windows")` for example output you might expect on the Windows cluster, which includes information about mapping of your paths onto those of the cluster, the version of R you will use, and other information.

If you have issues with `hipercow` we will always want to see the output of `hipercow_configuration()`.

# Running your first task

The first time you use the tools (ever, in a while, or on a new machine) we recommend sending off a tiny task to make sure that everything is working as expected:

```{r create}
id <- task_create_expr(sessionInfo())
```

This creates a new task that will run the expression `sessionInfo()` on the cluster. The `task_create_expr()` function works by so-called ["non standard evaluation"](https://adv-r.hadley.nz/metaprogramming.html) and the expression is not evaluated from your R session, but sent to run on another machine.

The `id` returned is just an ugly hex string:

```{r}
id
```

```{r, include = FALSE}
# Ensure that the task is finished, makes the rest of the doc nicer to read
task_wait(id)
```

Many other functions accept this `id` as an argument.  You can get the status of the task, which will have finished now because it really does not take very long:

```{r status}
task_status(id)
```

Once the task has completed you can inspect the result:

```{r result}
task_result(id)
```

Because we are using the "example" driver here, this is the same as the result that you'd get running `sessionInfo()` directly, just with more steps.  See `vignette("windows")` for an example that runs on Windows.

# Using functions you have written

It's unlikely that the code you want to run on the cluster is one of the functions built into R itself; more likely you have written a simulation or similar and you want to run *that* instead.  In order to do this, we need to tell the cluster where to find your code. There are two broad places where code that you want to run is likely to be found **script files** and **packages**; we start with the former here, and deal with packages in much more detail in `vignette("packages")`.

Suppose you have a file `simulation.R` containing some simulation:

```{r, results = "asis", echo = FALSE}
r_output(readLines("simulation.R"))
```

We can't run this on the cluster immediately, because the cluster does not know about the new function:

```{r create_walk1}
id <- task_create_expr(random_walk(0, 10))
task_wait(id)
task_status(id)
task_result(id)
```

(See `vignette("troubleshooting")` for more on failures.)

We need to tell hipercow to `source()` the file `simulation.R` before running the task. To do this we use `hipercow_environment_create()` to create an "environment" (not to be confused with R's environments) in which to run things:

```{r create_environment}
hipercow_environment_create(sources = "simulation.R")
```

Now we can run our simulation:

```{r create_walk2}
id <- task_create_expr(random_walk(0, 10))
task_wait(id)
task_result(id)
```

We have more to write on environments but briefly:

* You can have multiple environments and each task can be set to run in a different environment
* Each environment can source any number of source files, and load any number of packages
* This will become the mechanism by which environments on parallel workers (via `parallel`, `future` or `rrq`) will set up their environments

# Getting information about tasks

Once you have created (and submitted) tasks, they will be queued by the cluster and eventually run.  The hope is that we surface enough information to make it easy for you to see how things are going and what has gone wrong.

## Fetching information with `task_info()`

The primary function for fetching information about a task is `task_info()`:

```{r}
task_info(id)
```

This prints out core information about the task; its identifier (`r inline(id)`) and status (`success`), along with information about what sort of task it was, what expression it had, variables it used, the environment it executed in and the time that key events happened for the task (when it was created, started and finished).

This display is meant to be friendly; if you need to compute on this information, you can access the times by reading the `$times` element of the `task_info()` return value:

```{r}
task_info(id)$times
```

Likewise, the information about the task itself is within `$data`. To work with the underling data you might just unclass the object to see the structure:

```{r}
unclass(task_info(id))
```

but note that the exact structure is subject to (infrequent) change.

## Fetching logs with `task_log_show`

Every task will produce some logs, and these can be an important part of understanding what they did and why they went wrong.

You can view the log with `task_log_show()`

```{r}
task_log_show(id)
```

This prints the contents of the logs to the screen; you can access the values directly with `task_log_value(id)`.  The format of the logs will be generally the same for all tasks; after the header saying where we are running, some information about the task will be printed (its identifier, the time, details about the task itself), then any logs that come from calls to `message()` and `print()` within the queued function (within the "task logs" section; here that is empty because our task prints nothing).  Finally, a summary will be printed with the final status, final time (and elapsed time), then any warnings that were produced will be flushed (see `vignette("troubleshooting")` for more on warnings).

There is a second log too, the "outer" log, which is generally less interesting so it is not the default.  These logs come from the cluster scheduler itself and show the startup process that leads up to (and after) the code that hipercow itself runs.  It will differ from driver to driver.  In addition, this log may not be available forever; the windows cluster retains it only for a couple of weeks:

```{r}
task_log_show(id, outer = TRUE)
```

The logs returned by `task_log_show(id, outer = FALSE)` are the logs generated by the statement containing `Rscript -e`.

## Watching logs with `task_log_watch`

If your task is still running, you can stream logs to your computer using `task_log_watch()`; this will print new logs line-by-line as they arrive (with a delay of up to 1s by default). This can be useful while debugging something to give the illusion that you're running it locally.

Using `Ctrl-C` (or `ESC` in RStudio) to escape will only stop log streaming and not the underlying task.

# Parallel tasks

So far, the tasks we submitted have been run using a single core on the cluster, with no special other requests made. Here is a simple example using two cores; we'll use `hipercow_resources()` to specify we want two cores on the cluster, and `hipercow_parallel()` to say that we want to set up two processes on those cores, using the `parallel` package. (We also support [`future`](https://future.futureverse.org/)).

```{r parallel_task}
resources <- hipercow_resources(cores = 2)
id <- task_create_expr(
  parallel::clusterApply(NULL, 1:2, function(x) Sys.sleep(5)),
  parallel = hipercow_parallel("parallel"),
  resources = resources)
task_wait(id)
task_info(id)
```

Both of our parallel tasks are to sleep for 5 seconds. We use `task_info()` to report  how long it took for those two runs to execute; if they ran one-by-one, we'd expect around 10 seconds, but we are seeing a much shorter time than that, so our pair of processes are running at the same time.

For details on specifying resources and launching different kinds of parallel tasks, see `vignette("parallel")`.

# Understanding where variables come from

Suppose our simulation started not from 0, but from some point that we have computed locally (say `x`, imaginatively)

```{r}
x <- 100
```

You can use this value to start the simulation by running:

```{r create_walk3}
id <- task_create_expr(random_walk(x, 10))
```

Here the `x` value has come from the environment where the expression passed into `task_create_expr()` was found (specifically, we use the [`rlang` "tidy evaluation"](https://rlang.r-lib.org/reference/topic-defuse.html) framework you might be familiar with from `dplyr` and friends).

```{r result3}
task_wait(id)
task_result(id)
```

If you pass in an expression that references a value that does not exist locally, you will get a (hopefully) informative error message when the task is created:

```{r, error = TRUE, eval = FALSE}
id <- task_create_expr(random_walk(starting_point, 10))
#> Error in `rlang::env_get_list()`:
#> ! Can't find `starting_point` in environment.
```

# Other ways of creating tasks

Above, we've been using `task_create_expr` to create a task from an expression, and that has done some automatic analysis of our code to try and work out what we need to run the expression everywhere.  We hope this is useful, but the automatic-ness can make things a little "magical" and as a result as bit hard to reason about in some contexts.  We provide two other functions for creating tasks that are less magic: `task_create_explicit` and `task_create_call`.  Once you have created your task it should behave much the same way, and so the decision on what function to use should be primarily a question of which captures your intent best, or which you find easiest to understand.

## Less magic with `task_create_explicit`

Sometimes, the automatic variable detection algorithm will get confused:

```{r, error = TRUE}
mu <- 5
id <- task_create_expr(lapply(1:5, function(number) rnorm(number, mu)))
task_wait(id)
task_result(id)
```

Here, our automatic detection sees `number` in your code, decides that it's something that you need to evaluate the expression (because it can't see that it's an argument to the closure) and then fails to find `number` in the environment from which you called it.

The `task_create_explicit` has a simpler logic, accepting an expression only:

```{r}
id <- task_create_explicit(
  quote(lapply(1:5, function(number) rnorm(number, mu))),
  export = "mu")
task_wait(id)
task_result(id)
```

(RGF: I think that we're mixing together two things here and simply being able to override the analysis might be better...)

## No magic with `task_create_call`

The `task_create_call` interface is most similar to R's `do.call` or callr's `callr::r_bg`.  We take a function and a list of arguments and arrange to call your function with those arguments.

As an example of `do.call` for those not familiar, consider a function that computes roots of a quadratic equation:

```{r}
quadroot <- function(a, b, c) {
  -(b + c(-1, 1) * sqrt(b^2 - 4 * a * c)) / (2 * a)
}
```

We can call this as

```{r}
quadroot(1, 4, 3)
```
However, if we had a list of arguments like this:

```{r}
args <- list(a = 1, b = 4, c = 3)
```

we can call it using `do.call` as

```{r}
do.call(quadroot, args)
```

We can apply the same approach with hipercow:

```{r}
id <- task_create_call(quadroot, args)
task_wait(id)
task_result(id)
```

# Cancelling tasks

You can cancel a task if it has been submitted and not completed, using `task_cancel()`:

For example, here's a task that will sleep for 10 seconds, which we submit to the cluster:

```{r}
id <- task_create_expr(Sys.sleep(10))
```

Having decided that this is a silly idea, we can try and cancel it:

```{r}
task_cancel(id)
task_status(id)
task_info(id)
```

You can cancel a task that is submitted (waiting to be picked up by a cluster) or running (though not all drivers will support this; we need to add this to the example driver still, which will improve this example!).

You can cancel many tasks at once by passing a vector of identifiers at the same time.  Tasks that have finished (successfully or not) cannot be cancelled.

# Retrying tasks

There are lots of reasons why you might want to retry a task.  For example:

* it failed but you think it might work next time
* you updated a package that it used, and want to try again with the new version
* you don't like the output from some stochastic function and want to generate new output
* you cancelled the task but want to try again now

You can retry tasks with `task_retry()`, which is easier than submitting a new task with the same content, and also preserves a link between retried tasks.

Our random walk will give slightly different results each time we use it, so we demonstrate the idea with that:

```{r}
id1 <- task_create_expr(random_walk(0, 10))
task_wait(id1)
task_result(id1)
```

Here we ran a random walk and it got to `r tail(task_result(id1), 1)`, which is clearly not what we were expecting.  Let's try it again:

```{r}
id2 <- task_retry(id1)
```

Running `task_retry()` creates a *new* task, with a new id `r abbrev_id(id2)` compared with `r abbrev_id(id1)`.

Once this task has finished, we get a different result:

```{r}
task_wait(id2)
task_result(id2)
```

Much better!

We get a hint that this is a retried task from the `task_info()`

```{r}
task_info(id2)
```

You can see the full chain of retries here:

```{r}
task_info(id2)$retry_chain
```

Once a task has been retried it affects how you interact with the previous ids; by default they follow through to the most recent element in the chain:

```{r}
task_result(id1)
task_result(id2)
```

You can get the original result back by passing the argument `follow = FALSE`:

```{r}
task_result(id1, follow = FALSE)
task_result(id2)
```

Only tasks that have been completed (`success`, `failure` or `cancelled`) can be retried, and doing so adds a new task to the *end* of the chain; there is no branching. Retrying the `id1` here would create the chain `id1 -> id2 -> id3`, and following would select `id3` for any of the three tasks in the chain.

You cannot currently change any property of a retried task, we may change this in future.

```{r, include = FALSE}
cleanup()
```

# Running many tasks at once

Submitting a single task gets work off your laptop at least, but where a cluster really shines is in running many tasks at once.  We offer two functions for running groups of related tasks: `task_create_bulk_expr` and `task_create_bulk_call`.  These both achieve the same thing, but with interfaces that help you think about the problems in slightly different ways.

## The unreasonable effectiveness of `lapply`

The `lapply` function is one of the most useful functions that you will learn using R.  It's great because it abstracts away the details of a really common task.  Consider this `for` loop:

```r
results <- list()
for (i in seq_along(data)) {
  results[[i]] <- somefn(data[[i]])
}
```

This makes a little loop over some data, creating a list where we store the results of applying the function `somefn` to each piece of data.  But the majority of this is just "bookkeeping" and the simplicity of what we're tring to do is obscured.  So, a seasoned R user would probably prefer to write

```r
results <- lapply(data, somefn)
```

which we should all be able to agree is much simpler.  More than simple though, we've made a subtle change that is quite important: we are no longer dictating the order that the function is applied over - this frees us up to do things like run each of these functions in parallel (this is basis of `parallel::parLapply` and `furrr::pmap` among others), even on another machine.  A good rule of thumb for making the move from a for loop into a `lapply` that could theoretically be parallelised is "can I run my loop backwards and get the same answer".  In our case we probably can:

```r
results <- list()
for (i in rev(seq_along(data))) {
  results[[i]] <- somefn(data[[i]])
}
```

Reasons why you might not get the same answer for this block of code compared with the one that runs forwards might include:

* Your function uses random numbers
* Your function contains internal mutable state

We won't cover here how to cope with getting repeatable answers from functions that have these issues, but we will press on to show how to use hipercow to shift code that uses `lapply` to run in parallel using a cluster.

## `lapply`-like parallelism with `task_create_bulk_call`

Recall our quadratic root function `quadroot` above, which we used with `task_create_call`:

```{r}
id <- task_create_call(quadroot, list(a = 1, b = 4, c = 2))
task_wait(id)
task_result(id)
```

There are two ways that we can use `task_create_bulk_call` to submit many tasks that iterate over arguments to this function.  The first is very `lapply`-like where we loop over a vector of inputs. Suppose we want to vary `b` and compute roots, while keeping `a` and `c` unchanged from the above; to do this we can write:

```{r}
bs <- c(1, 2, 3, 4, 5, 6)
ids <- task_create_bulk_call(quadroot, bs, args = list(a = 1, c = 2))
hipercow_bundle_wait(ids)
hipercow_bundle_result(ids)
```

If you had a function with just one argument then you can be even simpler here and omit the `args` argument.  Note here that the vector we are looping over is used as the `b` argument following R's argument matching rules: because `a` and `c` are named they are matched first, leaving just `b` remaining to be matched as the first unnamed argument.  This is the conceptually the same as running:

```{r}
quadroot(2, a = 1, c = 2)
```

Now, suppose that we want to vary all three arguments among call (this is what you might do if running a simulation over a grid of possible parameters for example).  We can put all of the arguments to `quadroot` into a `data.frame`:

```{r}
abc <- expand.grid(a = 1, b = 1:4, c = 2:3)
abc
```

Then call `task_create_bulk_call` with this `data.frame` as the second argument:

```{r}
ids <- task_create_bulk_call(quadroot, abc)
hipercow_bundle_wait(ids)
hipercow_bundle_result(ids)
```

Here, we have called `quadroot` once per *row* in the `data.frame`, spreading the arguments `a`, `b` and `c` from each row for each call.

## Expression-based bulk calls with `task_create_bulk_expr`

You can achieve the same goal a different way using `task_create_bulk_expr`, which is more similar to `task_create_expr`, but uses the expression as a template into which it substitutes variables from your `data.frame`.  So we migfht write the above as:

```{r}
grp <- task_create_bulk_expr(quadroot(a, b, c), abc)
hipercow_bundle_wait(grp)
hipercow_bundle_result(grp)
```

Here, the variables `a`, `b` and `c` were swapped in, row-by-row from the data.frame `abc`.

## List columns

The above techniques end up using `data.frame`s for much of the management of tasks.  This is convenient as we can think as a row as the instructions for each task and columns as the inputs for each argument.  However, this is only familiar to most people if you have scalar parameters (i.e., in a `data.frame`, each column is a numeric, logical or character vector so that selecting the `i`th element is a scalar).  Sometimes though you need to pass in something longer than this (e.g., a vector of intervention times).  You can achieve this using *list columns*, which are frankly quite weird but can be very useful.  

A list column is a column where each element is an arbitrary R object so the column itself is a list.  They can be a nuisance to create:

```{r, error = TRUE}
b <- lapply(1:5, rnorm)
data.frame(a = 1:5, b = b)
```

You need to use `I()` to wrap the column:

```{r}
data.frame(a = 1:5, b = I(b))
```

Or, add them into an existing `data.frame`

```{r}
d <- data.frame(a = 1:5)
d$b <- b
d
```

Here's a function that carries out some simple calculations given `a` and `b`:

```{r}
fn <- function(a, b) {
  (var(b) / mean(b))^a
}
```

We can use the data.frame above with this function using the bulk submission functions `task_create_bulk_call`:

```{r}
ids <- task_create_bulk_call(fn, d)
hipercow_bundle_wait(ids)
hipercow_bundle_result(ids)
```

or with `task_create_bulk_expr`

```{r}
ids <- task_create_bulk_expr(fn(a, b), d)
hipercow_bundle_wait(ids)
hipercow_bundle_result(ids)
```

## Handling bundles

The above bits of code use `hipercow_bundle_wait` in place of `task_wait` and other bundle functions.  When you submit a bulk group of tasks, it is often useful to refer to them together as bundle; we often want to know if they are all complete, or get all the results.  When you get the results out, you will usually want them in the same order as the inputs.  The `hipercow_bundle_*` functions do this for you.

You will end up with far fewer bundles than tasks, and they will typically be more meaningful.  Because of this, we have a `hipercow_bundle_list` function that lists known bundles:

```{r}
hipercow_bundle_list()
```

You can then load any previous bundle with `hipercow_bundle_load`.  So if you keep track of the name of the bundle as you submit it, you can get back to it easily later.

If you are submitting groups of tasks that mean something to you, you may prefer to name your bundles:

```{r}
ids <- task_create_bulk_expr(fn(a, b), d, bundle_name = "final-runs")
```

then you get a name that you'll more easily remember rather than the randomly generated name.  In another session (on another machine, on another day, so long as you are in the same hipercow-enabled directory) you can then run:

```{r}
grp <- hipercow_bundle_load("final-runs")
hipercow_bundle_wait(grp)
hipercow_bundle_status(grp)
```

The main difference with the bundle functions is for retrying; `hipercow_bundle_retry` lets you filter the status for retrying, using the `if_status_in` argument, and that 