---
title: "Hello World"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Hello World}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
source("../vignettes/common.R")
vignette_root <- new_hipercow_root_path()
set_vignette_root(vignette_root)
```

```{r, echo = FALSE, results = "asis"}
add_header()
```

# Before starting

Do you have access to the cluster?  Try logging onto [the portal](https://mrcdata.dide.ic.ac.uk/hpc/index.php); if it reports that you don't have access then please request access by messaging Wes.  If you don't know your username and password then see the "Usernames and passwords" section in `vignette("details")`.

Ensure you have connected to the DIDE network by using ZScaler; see [instructions from ICT](https://www.imperial.ac.uk/admin-services/ict/self-service/connect-communicate/remote-access/unified-access/), or by being on a desktop PC or virtual machine within the building that is plugged in by Ethernet (this is less common now).

Install R and RStudio, if you have not already.

Install the packages by running

```r
install.packages(
  c("hipercow", "hipercow.windows"),
  repos = c("https://mrc-ide.r-universe.dev", "https://cloud.r-project.org"))
```

Map a network drive (see the "Paths and shares" section in `vignettes("details")`).

Start by creating an [RStudio project](https://support.posit.co/hc/en-us/articles/200526207-Using-RStudio-Projects) in your home directory, or in another network drive from above.  Once you have done this, you should be able to run

```{r}
getwd()
```

and recognise that this is a network path (see `vignette("details")` for more information if you are unclear on this).

# Hello world

## Very quickly

We recommend you do this before starting any cluster work:

```{r hello}
library(hipercow)
hipercow_init(driver = "windows")
hipercow_hello()
```

if this runs without error you're good to go.

## Still quite quickly

```{r, include = FALSE}
vignette_root <- new_hipercow_root_path()
set_vignette_root(vignette_root)
```

This is the state of my workstation:

```{r}
sessionInfo()
```

you can see that I run `r osVersion`, which might be different to you, but it's *certainly* different to the cluster. Our aim is to run this small piece of code on the cluster.

First, load the package

```{r}
library(hipercow)
```

Then create a hipercow root:

```{r initialising}
hipercow_init(driver = "windows")
```

Enqueue a task; here the expression `sessionInfo()`

```{r creating}
id <- task_create_expr(sessionInfo())
```

Then, wait for the task to return, followed by fetching its result:

```{r waiting}
task_wait(id)
task_result(id)
```

You can see that the `sessionInfo` result here is different to the above, as it ran on another machine.

# Next steps

* Start submitting more interesting tasks with `vignette("hipercow")`
* Learn how to install packages on the cluster in `vignette("packages")`
