---
title: "Packages and provisioning"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Packages and provisioning}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
library(hipercow)
source("common.R")
```

```{r, echo = FALSE, results = "asis"}
add_header()
```

Depending on where your packages come from and how specific you need to be on versioning, getting packages available for your packages on the cluster ranges from very straightforward to a constant struggle.

We install packages into a special directory within `hipercow/` by running a special job on the cluster itself using our [`conan2`](https://mrc-ide.github.io/conan2) package.  This aims to solve most of the common situations that we have seen, and has a couple of escape hatches that we hope will help with future problems.

There are three broad approaches to installing packages that we support:

* Dependency resolution and package installation using [`pkgdepends`](https://r-lib.github.io/pkgdepends) (`method = "pkgdepends"`)
* A bespoke script that you write yourself, and we run (`method = "script"`)
* Reproducible project environments with [`renv`](https://rstudio.github.io/renv/) (`method = "renv"`)

We expect that packages might come from any number of places:

* CRAN or Bioconductor
* A CRAN-like repository (such as an ["R universe"](https://mrc-ide.r-universe.dev) or a drat repository)
* A **public** GitHub repository, which one might ordinarily install via `remotes::install_github()`
* Local sources, available on your computer and in the hipercow directory

We don't currently support packages from private repositories, but let us know if that is a problem for you.

This all ends up being a lot of flexibility, so our (current) suggestion if you are starting out is to use the `pkgdepends` method with the `pkgdepends.txt` configuration a starting point as this will be the easiest for us to replicate issues with, and should end up being the most predicatable.

# Using `pkgdepends`

This is the method that is least prescriptive on your workflow and simplest.  We recommend this as a good starting place.

There are two places that we might get lists of packages to install:

* From a file `pkgdepends.txt` in the directory that also contains the `hipercow/` directory
* Manually, within a call to `hipercow_provision()`

We'll consider these in turn here.

## A list of packages

Create a file called `pkgdepends.txt` at the top of the hipercow root.  This file has a very simple format, with each line being the name of a package, or a [`pkgdepends` "package reference"](https://r-lib.github.io/pkgdepends/reference/pkg_refs.html), plus support for comments, blank lines and specifying additional repositories to install from.

So a complex case might look like

```
# Also use the ncov-ic universe
repo::https://ncov-ic.r-universe.dev

# Specific version of malaria simulation, via a tag
mrc-ide/malariasimulation@v1.6.0

# Package from CRAN
coda
```

* Blank lines are ignored
* Lines starting with `#` are comments and are also ignored
* Entries are separated by newlines
* Lines like `coda` represent packages from CRAN or another CRAN-like repository
* Lines like `mrc-ide/malariasimulation@v1.6.0` are a pkgdepends reference and can be quite complex to control what version is installed
* Lines like `repo::https://ncov-ic.r-universe.dev` add additional CRAN-like repositories (this is the recommended way of installing some of our packages like `odin` or `orderly2`)
* You can install packages that you have a `.tar.gz` for by adding a line like `local::./mypkg_0.1.0.tar.gz`

The [`pkgdepends` docs](https://r-lib.github.io/pkgdepends/reference/pkg_refs.html) provide lots of nice examples here.

```{r setup_pkgdepends, include = FALSE}
vignette_root <- new_hipercow_root_path()
hipercow::hipercow_init(vignette_root, driver = "windows")
writeLines("cowsay", file.path(vignette_root, "pkgdepends.txt"))
set_vignette_root(vignette_root)
```

We have an new empty hipercow root, containing a simple `pkgdepends.txt` file:

```{r, echo = FALSE}
dir_tree_hipercow(".")
```

The `pkgdepends.txt` file contains simply:

```{r, echo = FALSE, results = "asis"}
plain_output(readLines("pkgdepends.txt"))
```

We can provision this environment by running

```{r provision_pkgdepends}
hipercow_provision()
```

This launches a small task on the cluster and uses `pkgdepends` to install `cowsay` and all its dependencies.  If recent installations have been attempted, it will use cached copies, which should make things fairly snappy.

Once installed we can run a task that uses this package:

```{r cowsay}
id <- task_create_expr(cowsay::say("HiperMoo", "cow"))
task_wait(id)
task_log_show(id)
```

## Manually

Sometimes, regardless of how things are installed, you need to install something manually.  For example, you just want to change the version being used of some package, or `pkgdepends` fails to resolve a nice set of dependencies from some complex interdependent set of packages and you want to install a specific version on top of everything it did.

To do this, `hipercow_provision` with `method = "pkgdepends"` and `refs` as a character vector of package references, following the same format as `pkgdepends.txt`

```{r provision_manually}
hipercow_provision(method = "pkgdepends", refs = "cran::coda")
```

# Automatically, from an environment

The `pkgdepends` approach above duplicates some information already present in the packages listed in your calls to `hipercow_environment()`, implied in the files `source()`'d at the start of your task and from the local metadata about where your packages were installed from in the first place.

We try to poke around in your script files and find all the packages that you use, then we look at your installation, and from tha build a set of references that we hope will recreate the installation.

```{r setup_auto, include = FALSE}
vignette_root <- new_hipercow_root_path()
hipercow::hipercow_init(vignette_root, driver = "windows")
writeLines(c(
  "cowsay_fact <- function() {",
  '  cowsay::say("catfact", "cow")',
  "}"), file.path(vignette_root, "src.R"))
set_vignette_root(vignette_root)
```

We have an new empty hipercow root, this time containing some source file `src.R`:

```{r, echo = FALSE}
dir_tree_hipercow(".")
```

The file `src.R` contains the code that we want to run on the cluster:

```{r, echo = FALSE, results = "asis"}
r_output(readLines("src.R"))
```

We tell hipercow about this:

```{r create_env}
hipercow_environment_create(sources = "src.R")
```

Now, when we run `hipercow_provision()`, `hipercow` will check through our `src.R` file and seen that we use `cowsay`

```{r provision_auto1}
hipercow_provision()
```

Now our code works!

```{r use_auto}
id <- task_create_expr(cowsay_fact())
task_wait(id)
task_log_show(id)
```

If we had installed `cowsay` initially by running `remotes::install_github()` it would have noticed this and installed the package that way.  For example, the [`rfiglet` package](https://richfitz.github.io/rfiglet) is not available on CRAN and must be installed via `remotes`:

```r
remotes::install_github("richfitz/rfiglet")
```

```{r, include = FALSE}
local({
  code <- readLines("src.R")
  writeLines(c(
    code,
    "",
    "figlet_date <- function() {",
    "  print(rfiglet::figlet(as.character(Sys.Date())))",
    "}"), "src.R")
})
```

I've adjusted our `src.R` to use `rfiglet`:

```{r, echo = FALSE, results = "asis"}
r_output(readLines("src.R"))
```

and now when we call `hipercow_provision()` you can see it installs rfiglet too, automatically finding it on GitHub!

```{r provision_auto2}
hipercow_provision()
```

```{r use_auto2}
id <- task_create_expr(figlet_date())
task_wait(id)
task_log_show(id)
```


# Using a script

Sometimes you just need a lot more control than `pkgdepends` can provide and you want to do your own thing.  To support this, create a file `provision.R` in the root of your hipercow repository and within this write whatever you want to install packages.

```{r, include = FALSE}
vignette_root <- new_hipercow_root_path()
hipercow::hipercow_init(vignette_root, driver = "windows")
writeLines(c('install.packages("cowsay")',
             'cowsay::say("Moo", "cow")'),
           file.path(vignette_root, "provision.R"))
set_vignette_root(vignette_root)
```

Here, we have another empty hipercow root, containing a file `provision.R`:

```{r, echo = FALSE}
dir_tree_hipercow(".")
```

The `provision.R` file contains simply:

```{r, echo = FALSE, results = "asis"}
r_output(readLines("provision.R"))
```

Here, it's very simple, and just installs the package (plus runs a simple function from it).

```{r provision_script}
hipercow_provision()
```

Please be considerate and don't write scripts that will take very long to run (e.g., no longer than 10-15 minutes) or they may be killed by the cluster scheduler.  Please also let us know if you have a script that takes this long.

# Using renv

If you are using [`renv`](https://rstudio.github.io/renv/) to set up your packages, you can use this with hipercow, and this should be detected automatically.  With your renv project loaded and a lockfile created (so that `renv::status()` reports "No issues found -- the project is in a consistent state.") you should be able to run

```r
hipercow_provision()
```

and see `renv` build a new library based on your lockfile.  New tasks launched after this will use that library.

<!--
It would be nice to show renv usage here, but it's hard to get working reliably because we don't actually want renv running in the vignette process, and callr won't reliably pick up renv from a non-renv session!
-->

The workflow differs very slightly from using `renv` normally in that we take some steps to prevent `renv` updating the library automatically while your tasks run because this will result in disaster if multiple tasks trigger this in parallel.

# Some details about the process

The general hope is that relatively little configuration or specific calls are needed here, but this comes at the cost of a little magic.

If the `method` argument to `hermod_provision()` is not given, then we use a simple heuristic to select a method:

* If a file `provision.R` is present, we use the `script` method; this takes precedence because it's biggest, least clever hammer. You might use this as a last resort and so it should override any other method that you might be using.
* If a file `pkgdepends.txt` is present, we use the `pkgdepends` method; we think that this method is a reasonable tradeoff of predictability and usability and should cover most people's needs.
* Otherwise we'll try for an automatic installation; this is to help people get started quickly.

You can always pass the method explicitly and any arguments supported by that method (see `?hipercow_provision` for details).  For example, suppose you normally had things working with `pkgdepends` but hit trouble with some incompatible set of versions, you might write a file `fix_install.R` and run it with:

```r
hipercow_provision(method = "script", script = "fix_install.R")
```

If you interrupt the installation while it runs, it will not (currently) cancel the running task. We may change this in future.  Please don't submit another provisioning task until yours is finished (you can use the [hpc portal](https://mrcdata.dide.ic.ac.uk/hpc/index.php) to check easily).
