---
title: "Using stan"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using stan}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
source("../vignettes/common.R")
vignette_root <- new_hipercow_root_path(TRUE)
set_vignette_root(vignette_root)
```

```{r, echo = FALSE, results = "asis"}
add_header()
```

```{r, include = FALSE}
writeLines(c("repo::https://stan-dev.r-universe.dev", "cmdstanr"),
           "pkgdepends.txt")
```

[Stan](https://mc-stan.org/) is a platform for statistical modelling, using state-of-the art algorithms.  Unfortunately, it is very peculiar about how it needs things installed, and more unfortunately it plays poorly with other packages and they change their minds about how to install things every few versions.  If you have having trouble getting your compilers working with stan, rest assured you are in good company any have been so for a decade or so!

This vignette outlines some survival guides for using stan on the cluster.  They will likely stop working at some point in the future; please let us know when this happens and we can explore what the current recommended way of doing things has changed to, if there is one.

# Local installations

The `rstan` package is available on CRAN but has historically lagged releases that are available directly from the stan developers.

You can install `rstan` from CRAN but `cmdstanr` needs to be installed from the stan repo.  At the time of writing, here are links to the official documentation:

* [rstan](https://github.com/stan-dev/rstan/wiki/RStan-Getting-Started)
* [cmdstanr](https://mc-stan.org/cmdstanr/articles/cmdstanr.html)

You will need to install a suitable C++ toolchain.  Please refer to the stan docs above for details on this, and be aware that information found on StackOverflow or the stan discussion forums is often out of date.  You usually need a moderately recent version of R, too.

# CmdStan on the cluster

The `CmdStan` interface compiles stan programs into standalone executables, and somewhat isolates stan from the Rtools toolchain.  This is a newer approach than using `rstan`.  There are two parts here; the `cmdstanr` R package and the `CmdStan` executables.

## Installation and versions

The version of the package  that you use on your own machine is not important to these instructions (though it may be important to you depending on what model you are compiling).  On the cluster, you need to install a very recent version of `cmdstanr` (0.8.1 or newer, released on 6 June 2024), which is available from the stan repositories.

Your `pkgdepends.txt` should contain:

```{r, echo = FALSE, results = "asis"}
plain_output(readLines("pkgdepends.txt"))
```

(Be sure not to use the `https://mc-stan.org/r-packages` repo, still widely noted in documentation as that ships only older versions, all of which are broken.)

We currently support `CmdStan` version `2.35.0` - we would like to support multiple versions at once, but the internal gymnastics in `cmdstanr` make that quite hard to do.  Please let us know if you need a specific version different to whatever we have installed, and we will install it and update these instructions so that you can select it.

It is easy to make a mess on the cluster by following instructions from the stan developers which assume you are the only user of a machine, and that stan is the main piece of software you are interested in running.  This is especially the case if you follow instructions found on StackOverflow as these may refer to previous versions.  The fallout should be restricted to your own working space though.

Within a hipercow task, please do not run:

* `cmdstanr::install_cmdstan`: this will install over 1GB of small files onto the network shares.  If you need a specific version please let us know and we will make one available
* `cmdstanr::check_cmdstan_toolchain(fix = TRUE)`: this will do all sorts of damage, up to and including corrupting the Rtools installation that you need to install packages

## Compiling models

When you use `cmdstandr::cmdstan_model()` it will generate a file in the same directory as your `.stan` file.  This is fine when on your laptop, but is a poor choice in almost any other context, and especially when you might submit more than one job at a time (which, given you are using a cluster, you probably are).

You'll need to pass in a path as the `dir` to every call to `cmdstanr::cmdstan_model()` - unfortunately there's no way of doing this automatically.  So rather than

```r
mod <- cmdstanr::cmdstan_model("model.stan")
```

you can use

```r
mod <- cmdstanr::cmdstan_model("model.stan", dir = tempdir())
```

Using `tempdir()` guarantees that every process that starts up will be use a different directory, so you won't have to think about this very much.

The two issues being mitigated here are:

* if the toolchain used to compile stan models differs on your laptop and the cluster (this is true if you are using a different operating system, minor version of R, or if you have a different version of Rtools, or if you have a different version of cmdstanr) then the model that you have compiled on your machine will not work on the cluster, with perhaps difficult-to-diagnose errors.
* if you launch two or more tasks at once from the cluster, then every job will try and compile the model at the same time, and then try and write to the same file.  At most one of these tasks will succeed, with the others failing due to permissions issues (windows cannot write a file that is open).

The (potentially large) downside of using `tempdir()` is that every time you start a task that uses a stan program it will have to compile it.  Even for a simple program this is quite slow (say up to two minutes for a "hello world" type problem).  Before deciding if this is going to cause you issues, think about how long it will take to *run* your code.  If it's going to run for an hour, then this delay may be tolerable.

However, if your program is very fast once compiled, this might become a significant issue.  If you want to reuse the models then you need to do so in a way that (a) only hipercow tasks reuse the compiled models and (b) no task tries to recompile the model after it is created.  You might explore something like first submitting a task that runs

```r
mod <- cmdstanr::cmdstan_model("model.stan", dir = "models-hipercow")
```

(after creating the `models-hipercow` directory, as stan will error if it does not yet exist, and won't create it for you).  Then after that task completes successfully submitting a series of tasks that use the model.  These would include the same exact statement as above to load the model.

There appears to be no mechanism to either prevent compilation (you can pass `compile = FALSE` but the model is then unusable, even if it has been compiled), or to test to see if the model is up to date.  You can *probably* get away with this:

```r
cmdstan_model_but_dont_recompile <- function(path, dir = ".", ...) {
  path_exe <- file.path(dir, paste0(path, ".exe"))
  is_current <- file.exists(path_exe) &&
    file.info(path_exe)$mtime > file.info(path)$mtime
  if (!is_current) {
    stop(sprintf("stan model '%s' at '%s' is out of date!",
                 path, dir))
  }
  cmdstanr::cmdstan_model(path, exe_file = NULL, compile = TRUE,
                          dir = dir, ...)
}
```

which you can use in place of `cmdstanr::cmdstan_model`, though this will not behave as expected for models that include other files.  Good luck.

## A worked example

This example is designed partly so we can test everything works, but we may ask you to run this yourselves if you are having difficulty.

```{r, include = FALSE}
writeLines(
  c("run_stan <- function() {",
    "  path <- tempfile()",
    "  dir.create(path)",
    "  mod <- cmdstanr::cmdstan_model('code.stan', dir = path)",
    "  stan_data <- list(N = 10, y = c(0, 1, 0, 0, 0, 0, 0, 0, 0, 1))",
    "  fit_mcmc <- mod$sample(",
    "    data = stan_data,",
    "    seed = 123,",
    "    chains = 2)",
    "}"),
  "code.R")
writeLines(
  c("data {",
    "  int<lower=0> N;",
    "  array[N] int<lower=0, upper=1> y;",
    "}",
    "parameters {",
    "  real<lower=0, upper=1> theta;",
    "}",
    "model {",
    "  theta ~ beta(1, 5); // uniform prior on interval 0,1",
    "  y ~ bernoulli(theta);",
    "}"),
  "code.stan")
```

Start with a basic hipercow-for-windows setup:

```{r init}
library(hipercow)
hipercow_init(driver = "windows")
```

We have a simple stan model `code.stan`:

```{r, echo = FALSE, results = "asis"}
plain_output(readLines("code.stan"))
```

And a little wrapper around this, in `code.R`, which compiles the model and draws some samples:

```{r, echo = FALSE, results = "asis"}
r_output(readLines("code.R"))
```

We need to add our `code.R` into the default environment:

```{r}
hipercow_environment_create(sources = "code.R")
```

Our `provision.txt` is as above:

```{r, echo = FALSE, results = "asis"}
plain_output(readLines("pkgdepends.txt"))
```

Install everything with `hipercow_provision` as usual

```{r}
hipercow_provision()
```

Now we can run our stan task (we need a decent timeout here as compilation can be very slow):

```{r}
id <- task_create_expr(run_stan())
task_wait(id, timeout = 600)
```

Here are the logs from the stan task, which has hopefully worked for you, too.

```{r}
task_log_show(id)
```

Note that drawing samples from this took very little time (look at the last bit of stan output) compared with the total time taken for the task (see the last line of hipercow output).

Once compilation of models succeeds, all that remains is the much easier job of using your model to do amazing science.
